EP1162_Prelim_Data_Screening

### 2020-02-28 data-raw/ep1162_prelimDataScreening.Rmd
More data screening

---
title: "EP1162_Prelim_Data_Screening"
author: "Hardy Griesbauer"
date: "18/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)

@@

# Now edit .csv file manually

@@

@@

```{r}
dbhEdit2<-
  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))

@@

# Create a vector of years that are not to be interpolated
DoNotReplace<-  
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  filter(DBH=="DoNotReplace")

# Interpolation
dbhEdit3<-
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  mutate(DBH=as.numeric(DBH)) %>% 
  mutate(Year=paste("01-01",Year,sep="-")) %>% 
  mutate(Year=dmy(Year)) %>% 
  group_by(TreeID) %>% 
  pad() %>% 
  ungroup() %>% 
  mutate(Year=year(Year)) %>% 
  as.data.frame()

@@

# Run function by TreeID
dbhPred<-by(dbhEdit3, dbhEdit3[,"TreeID"],
   function(x) predict(lm(DBH ~ Year, data = x),new=data.frame(Year=c(1992,1994,1997,2009,2019))),simplify=F)

do

@@

>>>>

---
title: "EP1162_Prelim_Data_Screening"
author: "Hardy Griesbauer"
date: "18/02/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning = FALSE)
library(tidyverse)

@@

@@

dbhEdit2<-
  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))

@@

```{r}
# Load tree data with values to replace
load(here::here("data-raw","editedDBH.2020-02-27.RData"))

@@

# Create a vector of years that are not to be interpolated
DoNotReplace<-  
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  mutate(TreeID=factor(TreeID)) %>% 
  mutate(Year=as.numeric(Year)) %>% 
  filter(DBH=="DoNotReplace")

# Interpolation
dbhEdit3<-
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  mutate(DBH=as.numeric(DBH)) %>% 
  mutate(Year=paste("01-01",Year,sep="-")) %>% 
  mutate(Year=dmy(Year)) %>% 
  mutate(TreeID=factor(TreeID)) %>% 
  group_by(TreeID) %>% 
  pad() %>% 
  ungroup() %>% 
  mutate(Year=year(Year)) %>% 
  as.data.frame()

@@

# Run function by TreeID
dbhPred1<-by(dbhEdit3, dbhEdit3[,"TreeID"],
   function(x) predict(lm(DBH ~ Year, data = x),new=data.frame(Year=c(1992,1994,1997,2009,2019))),simplify=F)

# Bring datasets together 
dbhPred<-data.frame(TreeID=names(dbhPred1),do.call(rbind.data.frame,dbhPred1)) %>% # unpack output from by() function
  setNames(c("TreeID",c(1992,1994,1997,2009,2019))) %>% # rename columns
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to = "DBH.Pred") %>% # pivot longer 
  mutate(Year=as.numeric(Year)) %>%

  # Join with dbhEdit3
  left_join(dbhEdit3,by=c("TreeID","Year")) %>%

  # Join with doNot Replace
  left_join(DoNotReplace,by=c("TreeID","Year")) %>%

  # Formatting
  dplyr::select(Plot=Plot.x,TreeID,Year,DBH=DBH.x,DBH.Pred,DoNotRep=DBH.y) %>% 
  mutate_at(vars(DBH.Pred),round,1) %>% # round DBH.Pred

  # Find and replace
  mutate(DBH.New=DBH) %>% 
  mutate(DBH.New=coalesce(DBH.New,DBH.Pred)) %>% 
  mutate(DBH.New=replace(DBH.New,!is.na(DoNotRep),"NA")) %>%

  # Set new DBH column
  dplyr::select(Plot:Year,DBH=DBH.New) %>% 
  mutate(DBH.Pred=as.numeric(DBH)) %>% 
  dplyr::select(-DBH) %>% 
  mutate(Plot=factor(Plot)) %>% 
  mutate(Year=factor(Year)) %>% 
  mutate(TreeID=factor(TreeID))

@@

Now let's test to make sure that all DBH are now monotonic.

```{r}
library(MonoInc)

x<-
  dbhPred %>% 
  drop_na(DBH)

 mono.x<- by(x$DBH,x$TreeID,monotonic)

 # One tree still has decreasing DBH
 dbhPred %>% 
   filter(TreeID==as.numeric(names(which(mono.x==FALSE))))

One tree still has decreasing DBH. 1994 value for this tree is less than 1992.  This is because the linear function predicts values below the 1992 value.  Let's manually replace the 1994 value with interpolation between 1992 and 1997 values.

```{r}

  dbhPred[dbhPred$TreeID==31 & dbhPred$Year==1994,"DBH"]=22.5

*** 
## Now merge DBH values back into the dataset
```{r}
dat<-
  dat %>% 
  left_join(dbhPred,by=c("Plot","TreeID","Year")) %>% 
  arrange(Plot,TreeID,Year) %>% 
  mutate(DBH=coalesce(DBH.Pred,DBH)) %>% 
  mutate_at(vars(DBH),round,1) %>% 
  dplyr::select(-DBH.Pred)

# Heights
Now let's look at trees where height measurements aren't monotonically increasing.

*The chunk below isn't set to run, because we only want to do this once*

```{r eval=FALSE}
x<-
  dat %>% 
  drop_na(Height) %>% # drop rows with missing Height values 
  arrange(Plot,TreeID,Year) %>% 
  mutate(id=paste(Plot,TreeID,sep="."))

 mono.x<- by(x$Height,x$id,monotonic)

  # Trees with decreasing height measurements
 # write to .csv file
  x %>% 
   group_by(id) %>% 
   summarise(n=n()) %>% 
   filter(n>1) %>%  # filter tree records with more than one height measurement
   filter(id%in%names(which(mono.x==FALSE))) %>% 
   left_join(x,by="id") %>% 
   arrange(id) %>% 
   dplyr::select(Plot,TreeID,Species,Year,Height) %>% 
   pivot_wider(names_from="Year",values_from = "Height")

42 trees have decreasing DBH.  A visual inspection of the plot cards revealed several trees with broken tops (e.g., Plot 12 Tree 22, and Plot 18 Tree 346).

In almost all other cases, there is too much fluctuation to confidently delete height measurements for interpolation.  I recommend we leave the height measurements as is.
